---
title: "Backtest: predicting 2021 blind"
---

{{< include _ledger.md >}}

```{r}
source("_common.R")
have <- file.exists(pub("backtest_summary.csv"))
```

The same model, run as if it were 2021, using only what was known before that election: the 2014 national to 2016 local translation, the 2019 national vote and the 2016 local election. Who stood comes from the 2021 results themselves, just as the 2026 candidate lists supply it now. It never sees any 2021 total. The prediction is then scored against the IEC's official 2021 results.

```{r}
#| results: asis
if (!have) {
  cat("::: {.callout-note}\n## Not run yet\nThe backtest needs the 2014 national results (a browser download) and the 2016 local results. Until both are present it is skipped, and the forecast runs without it.\n:::\n")
} else {
  s <- rd("backtest_summary.csv")
  cat(sprintf("**Run:** %s.\n\n", s$label[1]))
}
```

## Against the simple rules

A forecast model is only worth its complexity if it beats simple rules: repeat the previous local election, or use the latest national vote as it stands. Every forecast and rule is scored on the same cases: every party on each council's ballot (see [corrections](corrections.qmd), E1).

```{r}
if (have) rd("backtest_versus_rules.csv") |>
  transmute(Rule = rule, `Seat error (mean abs, per party per council)` = round(seat_mae, 2),
            `Council control correct` = pct(control_correct), `Ward winners correct` = pct(ward_correct)) |>
  knitr::kable()
```

## Is the uncertainty honest?

A calibrated forecast's 90% intervals contain the actual result about 90% of the time. Far fewer means over-confidence; far more means vagueness.

```{r}
if (have) rd("backtest_summary.csv") |>
  transmute(Spreads = spreads, `Seat intervals` = seat_intervals, `50% intervals hit` = pct(coverage50),
            `90% intervals hit` = pct(coverage90), `Seat CRPS (lower is better)` = round(seat_crps, 3),
            `Ward log score (higher is better)` = round(ward_log_score, 3),
            `Seats won by parties the model could not see` = new_entrant_seats) |>
  knitr::kable()
```

```{r}
#| fig-height: 3
if (have) {
  rd("backtest_seats.csv") |>
    ggplot(aes(pit)) + geom_histogram(breaks = seq(0, 1, 0.1), fill = "#9FB3BC", colour = "#F3F5F4") +
    geom_hline(yintercept = nrow(rd("backtest_seats.csv")) / 10, linetype = "22", colour = "#1E2B33") +
    labs(x = "Where the actual seat count fell in the forecast distribution (PIT)", y = "Party x council cases") +
    theme_lge()
}
```

If the uncertainty is right, the bars are level with the dashed line. A U shape means results too often fell in the tails (over-confident); a hump in the middle means intervals were too wide.

## Estimating the political weather

The two spreads in assumption A02 were judgement. The backtest measures them: each party's council-level surprise is split into a part shared across the province and a part specific to each council.

```{r}
if (have) rd("backtest_shock_sds.csv") |>
  transmute(Parameter = parameter, `Before (judgement)` = round(config, 3), `Estimated` = round(estimated, 3),
            `Based on` = based_on) |>
  knitr::kable()
```

With a single election, the province-wide spread rests on only a handful of parties, so it is a rough estimate. The council-level spread rests on many more cases.

## Choosing the settings: a pre-registered experiment

The backtest compared 24 settings: learned local premiums kept or dropped, crossed with four province-wide and three council-level spreads. The rule was fixed in the [model log](log.qmd) before any result was seen: choose the lowest seat CRPS among settings whose 90% coverage lies between 80% and 97%.

```{r}
if (file.exists(pub("backtest_grid.csv"))) rd("backtest_grid.csv") |>
  arrange(seat_crps) |>
  transmute(Premiums = intercepts, Province = province, Council = muni,
            `Seat CRPS` = round(seat_crps, 3), `90% coverage (PIT)` = pct(pit_cov90),
            `Ward log score` = round(ward_log_score, 3), `Control Brier` = round(control_brier, 3),
            `Seat error` = round(seat_mae_median, 2),
            Result = case_when(chosen ~ "chosen", !eligible ~ "outside coverage band", TRUE ~ "")) |>
  knitr::kable()
```

## New parties: a pre-registered test

Parties with no history won seats the model could not see. The newcomer module was tested by running the 2021 backtest with and without it, using a prior built only from 2016's newcomers. The rule, fixed in the [model log](log.qmd) before the result: adopt the module if seat CRPS is lower with it and 90% coverage stays between 80% and 97%.

```{r}
if (file.exists(pub("backtest_newcomers.csv"))) {
  x <- rd("backtest_newcomers.csv")
  if ("module" %in% names(x)) x |>
    transmute(Run = module, `Newcomer cases` = newcomer_cases, `Seat CRPS` = round(seat_crps, 3),
              `90% coverage (PIT)` = pct(pit_cov90), `Seat error` = round(seat_mae_median, 2),
              `Council control correct` = pct(control_modal_correct), `Ward winners correct` = pct(ward_modal_correct),
              `Seats won by unseen parties` = new_entrant_seats,
              `Paired CRPS difference (90% interval)` = if_else(is.na(crps_diff), "",
                sprintf("%.3f (%.3f to %.3f)", crps_diff, crps_diff_lo, crps_diff_hi)),
              Decision = if_else(module == "with newcomers", if_else(adopt, "adopt", "do not adopt"), "")) |>
    knitr::kable()
  else cat(x$note[1])
}
```

## Ward winners: a pre-registered experiment

In the 2021 backtest the model picked fewer ward winners correctly than simply using the latest national vote. Twelve variants were compared, with the rule fixed in the [model log](log.qmd) before the result: slopes as fitted, corrected for noise in the 2019 shares, or fixed at 1; the ward/PR ticket-splitting ratio on or off; learned premiums as fitted or halved. The winner has the lowest ward Brier score among variants that are no more than 0.005 worse on seat CRPS and stay calibrated. It is adopted only if it beats the current model.

```{r}
if (file.exists(pub("backtest_ward_grid.csv"))) rd("backtest_ward_grid.csv") |>
  arrange(ward_brier) |>
  transmute(Slopes = b_mode, `Ticket-split ratio` = if_else(ward_ratio, "on", "off"), Premiums = if_else(shrink == 1, "as fitted", "halved"),
            `Ward Brier` = round(ward_brier, 4), `Ward winners correct` = pct(ward_modal_correct, 1),
            `Seat CRPS` = round(seat_crps, 4), `90% coverage` = pct(pit_cov90),
            Result = case_when(adopt ~ "adopted", current ~ "current model", best ~ "best, not better than current",
                               !eligible ~ "worse on seats or calibration", TRUE ~ "")) |>
  knitr::kable()
```

```{r}
if (file.exists(pub("backtest_wards.csv"))) {
  w <- rd("backtest_wards.csv")
  w |> count(case, name = "Wards") |> rename(Outcome = case) |> knitr::kable()
}
```

## Premium basis: a pre-registered experiment

A party's learned local premium compares its local vote with its national vote. Measured against the *earlier* national election, the premium also absorbs whatever swing happened between the two elections (O10). The alternative measures it against the national vote *interpolated* to the local election's date. Four variants were compared, with the rule fixed in the [model log](log.qmd) before the result: the lowest seat CRPS among calibrated variants that are no more than 0.005 worse on ward Brier score, adopted only if it beats the current model. The interpolated basis at half strength won and is now the current model (L045). It corrected the DA's level but not the ANC's, and it cost two councils on control, which the rule did not consider. The interval column, added after the decision, shows how far each variant is from the current model, with a bootstrap over councils.

```{r}
pa <- \(p, a) if_else(is.na(p) | is.na(a), "n/a", sprintf("%s / %s", pct(p, 1), pct(a, 1)))
if (file.exists(pub("backtest_premium_grid.csv"))) {
  pg <- rd("backtest_premium_grid.csv")
  if (!"crps_diff" %in% names(pg)) pg <- mutate(pg, crps_diff = NA_real_, crps_diff_lo = NA_real_, crps_diff_hi = NA_real_)
  pg |>
  arrange(seat_crps) |>
  transmute(`Premium basis` = premium_basis, Strength = if_else(shrink == 1, "full", "halved"),
            `Seat CRPS` = round(seat_crps, 4),
            `vs current (90% interval)` = if_else(current | is.na(crps_diff), "",
                                                   sprintf("%+.4f (%+.4f to %+.4f)", crps_diff, crps_diff_lo, crps_diff_hi)),
            `90% coverage` = pct(pit_cov90), `Ward Brier` = round(ward_brier, 4),
            `Councils correct` = pct(control_modal_correct),
            `DA 2021: predicted / actual` = pa(da_pred, da_actual),
            `ANC 2021: predicted / actual` = pa(anc_pred, anc_actual),
            Result = case_when(adopt ~ "adopted", current ~ "current model", best ~ "best, not better than current",
                               !eligible ~ "fails calibration or ward condition", TRUE ~ "")) |>
  knitr::kable()
}
```

## Noise centring: a pre-registered experiment

The simulation adds ward and voting-district noise to every party with one spread, on the log scale. Through the conversion to shares, that noise lowers a dominant party's *average* share even though each party's noise is centred on zero: a party on 85% of a district loses share on average. Whether that is a bias depends on what the fitted translation predicts. Two tests decide, with the rule fixed in the [model log](log.qmd) (L047) before the result: the 2021 backtest with and without mean-preserving noise, and a closure test asking whether each fitted translation reproduces the local election it was fitted on.

```{r}
#| results: asis
if (file.exists(pub("noise_decision.csv"))) {
  nd <- rd("noise_decision.csv")
  cat(sprintf("**Result:** %s. Seat CRPS difference (%s minus %s) %+.4f, 90%% interval %+.4f to %+.4f. Closure error: %s %.1f points, %s %.1f points.\n\n",
              nd$verdict, nd$alternative, nd$current, nd$crps_diff, nd$crps_diff_lo, nd$crps_diff_hi,
              nd$current, 100 * nd$closure_mae_current, nd$alternative, 100 * nd$closure_mae_alternative))
}
if (file.exists(pub("backtest_noise_grid.csv"))) {
  rd("backtest_noise_grid.csv") |>
    transmute(`Noise` = if_else(noise_centring == "mean", "mean-preserving", "centred on log share"),
              `Seat CRPS` = round(seat_crps, 4), `90% coverage` = pct(pit_cov90), `Ward Brier` = round(ward_brier, 4),
              `Control Brier` = round(control_brier, 3),
              `DA 2021: predicted / actual` = pa(da_pred, da_actual),
              `ANC 2021: predicted / actual` = pa(anc_pred, anc_actual),
              Current = if_else(current, "yes", "")) |>
    knitr::kable() |> print()
}
if (file.exists(pub("closure.csv"))) {
  cat("\n*Closure test: each translation predicting the local election it was fitted on (Western Cape, the parties it was fitted for).*\n\n")
  rd("closure.csv") |> filter(scope == "Western Cape") |> group_by(cycle) |> slice_max(actual, n = 4) |> ungroup() |>
    transmute(Cycle = cycle, Party = party, Actual = pct(actual, 1), `Deterministic (mean-preserving)` = pct(deterministic, 1),
              `With noise (current)` = pct(noisy, 1)) |>
    knitr::kable() |> print()
}
```

### What the kept choice means for 2026

The test kept the current noise. The closure test nonetheless shows it pulls a dominant party's share down in-sample, by several points in Cape Town, while its out-of-sample advantage in 2021 came from offsetting errors (MODEL-LOG L048). This table re-runs the 2026 forecast with mean-preserving noise. It is a sensitivity, not a forecast: the gap between the two columns is a measure of how much the headline rests on this unresolved choice.

```{r}
#| results: asis
if (file.exists(pub("noise_sensitivity.csv"))) {
  ns <- rd("noise_sensitivity.csv")
  vs <- ns |> filter(measure == "PR vote share (median)", muni_code %in% c("Western Cape", "CPT")) |>
    group_by(muni_code) |> slice_max(forecast, n = 5) |> ungroup() |>
    transmute(Area = if_else(muni_code == "CPT", "Cape Town", muni_code), Party = outcome,
              Forecast = pct(forecast, 1), `Mean-preserving noise` = pct(alternative, 1),
              Difference = sprintf("%+.1f points", 100 * (alternative - forecast)))
  print(knitr::kable(vs))
  ctl <- ns |> filter(measure == "control probability") |>
    filter(abs(alternative - forecast) >= 0.10) |> arrange(desc(abs(alternative - forecast))) |>
    transmute(Council = muni_code, Outcome = outcome, Forecast = pct(forecast), `Mean-preserving noise` = pct(alternative))
  if (nrow(ctl)) { cat("\n*Control outcomes whose probability moves by 10 points or more:*\n\n"); print(knitr::kable(ctl)) }
}
```

## Calibration and spreads: a pre-registered experiment

The noise experiment above exposed two defects that offset each other in 2021 ([model log](log.qmd) L048). Predicting its own election, the translation overstated the DA and understated the ANC (O13). One pooled noise spread, dominated by small parties' sampling noise, pulled the DA down (O12). Five variants fix these singly and together. Each is scored on the 2021 backtest and on whether it reproduces the elections it was fitted on. The rule was fixed in L049 before any result: a variant must reproduce its own elections to within a point, stay calibrated, lose no more than 0.005 on ward Brier and 0.02 on council control, and not be demonstrably worse on seats. The lowest seat CRPS among qualifying variants is adopted.

```{r}
#| results: asis
if (file.exists(pub("calibration_decision.csv"))) {
  cd <- rd("calibration_decision.csv")
  cat(if (any(cd$adopt)) sprintf("**Result:** variant %s adopted.\n\n", cd$variant[cd$adopt]) else
    "**Result:** no variant qualified, so the model is unchanged and frozen for 2026 (L050). Variant D came closest, with the best ward Brier of any model tested, but missed the closure threshold and the control guard narrowly; the thresholds were not relaxed after the fact.\n\n")
  cd |>
    transmute(Variant = paste(variant, description, sep = ": "), `Seat CRPS` = round(seat_crps, 4),
              `vs current (90% interval)` = if_else(current, "", sprintf("%+.4f (%+.4f to %+.4f)", crps_diff, crps_diff_lo, crps_diff_hi)),
              `Closure error` = sprintf("%.1f pts", 100 * closure_mae), `90% coverage` = pct(pit_cov90),
              `Ward Brier` = round(ward_brier, 4), `Control Brier` = round(control_brier, 3),
              `DA 2021` = pa(da_pred, da_actual), `ANC 2021` = pa(anc_pred, anc_actual),
              Result = case_when(adopt ~ "adopted", current ~ "current model", qualifies ~ "qualifies",
                                 !closure_ok ~ "fails closure", !seats_ok ~ "worse on seats",
                                 !coverage_ok ~ "fails coverage", !ward_ok ~ "fails ward guard",
                                 !control_ok ~ "fails control guard", TRUE ~ "")) |>
    knitr::kable() |> print()
}
```

## What the model could not see

```{r}
if (have) {
  b <- rd("backtest_blind.csv")
  if (nrow(b)) b |> transmute(Council = muni_code, Party = party, `Seats won` = actual) |> knitr::kable()
  else cat("Every party that won seats was visible to the model.")
}
```
