---
title: "Data and model checks"
---

{{< include _ledger.md >}}

Run automatically on every build. A check reports what it found and how much it affects; it never silently repairs anything. Failures are shown here and on the affected municipality pages.

```{r}
source("_common.R")
rd("checks.csv") |>
  mutate(Status = sprintf('<span class="status-%s">%s</span>', status, status)) |>
  transmute(ID = id, Area = area, Check = description, Affected = n_affected, Detail = detail,
            Severity = severity, Status) |>
  knitr::kable(escape = FALSE)
```

## Human confirmations

Some things only local knowledge can settle, such as whether a party that no longer appears on a ballot has folded or changed its name. Those judgements are recorded here with their evidence. A disputed item stays flagged until its sources agree.

```{r}
if (file.exists(pub("party_status.csv"))) {
  rd("party_status.csv") |>
    transmute(Council = muni_code, Party = party, Status = status, `2026 name` = name_2026,
              Evidence = evidence, Source = source, Date = date) |>
    knitr::kable()
} else {
  cat("No confirmations recorded yet.")
}
```
