```{r}
source("../_common.R")
m_meta <- council_meta |> filter(muni_code == muni)
m_info <- municipalities |> filter(muni_code == muni)
m_seats <- seat_summary |> filter(muni_code == muni)
m_ctl <- control |> filter(muni_code == muni)
p_hung <- sum(m_ctl$prob[m_ctl$outcome == "No majority"])
top <- m_seats |> slice_max(p_largest, n = 1, with_ties = FALSE)
cols <- party_colours(m_seats$party)
unverified <- str_detect(coalesce(m_info$seats_source, ""), "DERIVED")
```

::: {.lead}
No party wins a majority in `r pct_clip(p_hung)` of simulations. `r top$party` is the most likely largest party (`r pct_clip(top$p_largest)`). A majority needs `r m_meta$majority` of `r m_meta$total_seats` seats.
:::

```{r}
#| results: asis
if (unverified) cat("::: {.callout-warning}\n## Council size not yet verified\nThe", m_meta$total_seats,
                    "seats used here are derived from the ward count, not taken from the MEC's notice. Seat numbers on this page may shift when it is confirmed (check C07).\n:::\n")
```

## Seats

```{r}
#| fig-height: !expr max(2.5, 0.42 * nrow(m_seats) + 1)
m_seats |>
  mutate(party = fct_reorder(party, seats_mean)) |>
  ggplot(aes(y = party, colour = party)) +
  geom_vline(xintercept = m_meta$majority, linetype = "22", colour = "#1E2B33", linewidth = 0.4) +
  annotate("text", x = m_meta$majority, y = Inf, label = paste("majority:", m_meta$majority),
           vjust = 1.4, hjust = 1.08, size = 3.2, colour = "#1E2B33") +
  geom_segment(aes(x = seats_q05, xend = seats_q95, yend = party), linewidth = 3.2, alpha = 0.35,
               lineend = "round") +
  geom_point(aes(x = seats_median), size = 3.2) +
  scale_colour_manual(values = cols, guide = "none") +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0.02, 0.06))) +
  labs(x = "Seats: median, with the range covering 90% of simulations", y = NULL) +
  theme_lge()
```

```{r}
#| fig-height: 3.6
top_parties <- m_seats |> slice_max(seats_mean, n = 6) |> pull(party)
seat_dist |>
  filter(muni_code == muni, party %in% top_parties) |>
  mutate(party = factor(party, levels = top_parties)) |>
  ggplot(aes(seats, prob, fill = party)) +
  geom_col(width = 0.85) +
  facet_wrap(~party, nrow = 2, scales = "free_x") +
  scale_fill_manual(values = cols, guide = "none") +
  scale_y_continuous(labels = \(x) pct(x)) +
  labs(x = "Seats", y = "Share of simulations") +
  theme_lge() + theme(panel.grid.major.y = element_line(colour = "#CCD5D2", linewidth = 0.3))
```

## Coalition arithmetic

These are the smallest combinations that reach a majority, and how often each is available. Arithmetic only: whether parties would actually govern together is politics, and not modelled.

```{r}
coalitions |>
  filter(muni_code == muni) |>
  slice_max(p_minimal_winning, n = 8, with_ties = FALSE) |>
  transmute(Coalition = coalition, `Available in` = pct_clip(p_minimal_winning),
            `Median seats` = median_seats) |>
  knitr::kable()
```

## Wards

```{r}
#| fig-height: 5
lead <- ward_probs |> filter(muni_code == muni) |> slice_max(p_win, n = 1, by = ward_id, with_ties = FALSE)
g <- wards_geo |> filter(muni_code == muni) |> left_join(lead, by = c("muni_code", "ward_id"))
ggplot(g) +
  geom_sf(aes(fill = party, alpha = p_win), colour = "#F3F5F4", linewidth = 0.2) +
  scale_fill_manual(values = cols, name = "Most likely winner") +
  scale_alpha_continuous(range = c(0.25, 1), limits = c(0.3, 1), guide = "none") +
  theme_void() + theme(legend.position = "right", plot.background = element_rect(fill = "#F3F5F4", colour = NA),
                       legend.text = element_text(size = 9), legend.title = element_text(size = 10))
```

Paler wards are closer contests.

```{r}
lead |>
  arrange(p_win) |>
  transmute(Ward = ward_id, `Most likely winner` = party, Probability = pct_clip(p_win)) |>
  knitr::kable()
```
